Alternative Methods for Implementing Explicit and Finding Implicit Sharing in embedded DSLs

### 2022-12-16 paper/paper.tex
benchmarking toy problem

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
\lstset{language=haskell}
%
\title{Alternative Methods for Implementing
    Explicit and Finding Implicit Sharing in embedded DSLs}
%
\titlerunning{Alternative Explicit and Implicit Sharing}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Curtis D'Alves \and
Lucas Dutton \and
Steven Gonder \and
Christopher Kumar Anand
}
%
\authorrunning{C. D'Alves et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{McMaster University, 1280 Main St W Hamilton, Canada}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
TODO The abstract should briefly summarize the contents of the paper in
150--250 words.

@@

instance Exp Graph where
  constant x = Graph (hashcons $ NConstant x)
  variable x = Graph (hashcons $ NVariable x)
  add e1 e2 = Graph (do
                     h1 <- unGraph e1
                     h2 <- unGraph e2
                     hashcons $ NAdd h1 h2)
\end{minted}
The trick to uncovering sharing is in the
\mintinline{hashcons} function, which inserts a new node into the current DAG, but not
before checking if it is already there.
\begin{minted}{haskell}
hashcons :: Node -> State DAG NodeID
hashcons e = do
  DAG m <- get
  case lookup_key e m of
    Nothing -> let (k,m') = insert e m
               in put (DAG m') >> return k
    Just k -> return k
\end{minted}
The technique is essentially that of
hash-consing, popularized by its use in LISP compilers, but discovered by Ershov in 1958 \cite{ershov1958:consing}. Other works have explored the use
of type-safe hash-consing in embedded DSLs, see \cite{filliatre:typesafeconsing}.

@@

Our proposed solution is to build a serialized AST using byte strings for each node along with our
DAG.
The byte string stays outside monad, while the DAG remains inside.
We can do this efficiently by replacing the \mintinline{haskell}{BiMap}
with a trie.
In our toy example, we use package \texttt{bytestring-trie}.
% TODO reference literature on tries

@@

data DAG = DAG { unTrie :: Trie (Node,NodeID)
               , maxID  :: NodeID
               } deriving Show
\end{minted}
This looks a bit different because the \mintinline{haskel}{BiMap}
was a bijective relation between nodes and node ids,
whereas the trie maps byte strings to pairs (node,node id).
To get the DAG expressed as a relation, project out the values of the trie.

To prevent confusion, we rename the hash-consing function to \mintinline{haskel}{triecons}, as follows
\begin{minted}{haskell}
triecons :: ByteString -> Node -> State DAG NodeID
triecons sAST node = do
  DAG trie maxID <- get
  case Trie.lookup sAST trie of
    Nothing -> let maxID' = maxID+1
                   trie' = Trie.insert sAST (node,maxID') trie
                in do put $ DAG trie' maxID'
                      return maxID'
    Just (_,nodeID) -> return nodeID
\end{minted}
We use it to implement the DAG-building instance of the DSL,
which looks a lot like the previous instance.
The substantial differences are the 
\mintinline{haskell}{buildStringAST} calls
which you can think of as pretty printing, but optimized for the trie,
and the use of \mintinline{haskell}{seqArgs} (explained below):
\begin{minted}{haskell}
instance Exp Graph where
  constant x = let
    node = NConstant x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NConstant x) sAST
  variable x = let
    node = NVariable x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NVariable x) sAST
  add e1 e2 = let
      sAST = buildStringAST "nadd" [e1,e2]
      sT = do ns <- seqArgs [e1,e2]
              case ns of
                [n1,n2] -> triecons sAST $ NAdd n1 n2
                _ -> error "black magic"
    in Graph sT sAST
\end{minted}
The magic is in \mintinline{haskell}{seqArgs}.
We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
fail to look up its corresponding byte string AST in the Trie.
\begin{minted}{haskell}
seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST) =
      do DAG trie _ <- get
         case Trie.lookup sAST  trie of
           Nothing -> sT
           Just (_,nodeID) -> return nodeID
  in sequence $ map seqArg inps
\end{minted}

>>>>

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
\lstset{language=haskell}
%
\title{Alternative Methods for Implementing
    Explicit and Finding Implicit Sharing in embedded DSLs}
%
\titlerunning{Alternative Explicit and Implicit Sharing}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Curtis D'Alves \and
Christopher Kumar Anand \and
Lucas Dutton \and
Steven Gonder
}
%
\authorrunning{C. D'Alves et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{McMaster University, 1280 Main St W Hamilton, Canada}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
TODO The abstract should briefly summarize the contents of the paper in
150--250 words.

@@

instance Exp Graph where
  constant x = Graph (hashcons $ NConstant x)
  variable x = Graph (hashcons $ NVariable x)
  add e1 e2 = Graph (do
                     h1 <- unGraph e1
                     h2 <- unGraph e2
                     hashcons $ NAdd h1 h2)
\end{minted}
The trick to uncovering sharing is in the
\mintinline{haskell}{hashcons} function, which inserts a new node into the current DAG, but not
before checking if it is already there.
\begin{minted}{haskell}
hashcons :: Node -> State DAG NodeID
hashcons e = do
  DAG m <- get
  case lookup_key e m of
    Nothing -> let (k,m') = insert e m
               in put (DAG m') >> return k
    Just k -> return k
\end{minted}
The technique is essentially that of
hash-consing, popularized by its use in LISP compilers, but discovered by Ershov in 1958 \cite{ershov1958:consing}. Other works have explored the use
of type-safe hash-consing in embedded DSLs, see \cite{filliatre:typesafeconsing}.

@@

Our proposed solution is to build a serialized AST using byte strings for each node along with our
DAG.
The byte string stays outside monad, while the DAG remains inside.
We can do this efficiently by replacing the \mintinline{haskell}{BiMap}
with a trie.
In our toy example, we use the package \texttt{bytestring-trie}.
% TODO reference literature on tries

@@

data DAG = DAG { unTrie :: Trie (Node,NodeID)
               , maxID  :: NodeID
               } deriving Show
\end{minted}
This looks a bit different because the \mintinline{haskell}{BiMap}
was a bijective relation between nodes and node ids,
whereas the trie maps byte strings to pairs (node,node id).
To get the DAG expressed as a relation, project out the values of the trie.

To prevent confusion, we rename the hash-consing function to \mintinline{haskell}{triecons}, as follows
\begin{minted}{haskell}
triecons :: ByteString -> Node -> State DAG NodeID
triecons sAST node = do
  DAG trie maxID <- get
  case Trie.lookup sAST trie of
    Nothing -> let maxID' = maxID+1
                   trie' = Trie.insert sAST (node,maxID') trie
                in do put $ DAG trie' maxID'
                      return maxID'
    Just (_,nodeID) -> return nodeID
\end{minted}
We use it to implement the DAG-building instance of the DSL,
which looks a lot like the previous instance.
The substantial differences are the 
\mintinline{haskell}{buildStringAST} calls
which you can think of as pretty printing, but optimized for the trie,
and the use of \mintinline{haskell}{seqArgs} (explained below):
\begin{minted}{haskell}
instance Exp Graph where
  constant x = let
    node = NConstant x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NConstant x) sAST
  variable x = let
    node = NVariable x
    sAST = buildStringAST node []
    in Graph (triecons sAST $ NVariable x) sAST
  add e1 e2 = let
      sAST = buildStringAST "nadd" [e1,e2]
      sT = do ns <- seqArgs [e1,e2]
              case ns of
                [n1,n2] -> triecons sAST $ NAdd n1 n2
                _ -> error "black magic"
    in Graph sT sAST
\end{minted}
The magic is in \mintinline{haskell}{seqArgs}.
We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
fail to look up its corresponding byte string AST in the Trie.
\begin{minted}{haskell}
seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST) =
      do DAG trie _ <- get
         case Trie.lookup sAST  trie of
           Nothing -> sT
           Just (_,nodeID) -> return nodeID
  in sequence $ map seqArg inps
\end{minted}
